Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-04 paper/main.tex
more work on POPL version

%Comment for anonymization
%\def\fullversion{}

\documentclass[nocopyrightspace]{sigplanconf}

@@

 % \usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
 %  \DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
 %  \captionsetup[figure]{format=myformat}
 % \renewcommand{\captionlabelfont}{\bf}

@@

\input{definitions}

\begin{document}

% \abovedisplayskip=4pt
% \belowdisplayskip=4pt
% \abovedisplayshortskip=4pt
% \belowdisplayshortskip=4pt

@@

In this section, we informally introduce the quantitative program
logic and the automatic amortized analysis for for Clight programs.
The purpose of this section is to build some intuition that should
make the following sections easy to read.

@@

To obtain a compositional analysis we also have to take into account
the state resulting from the program execution.  We thus use two
potential functions, one that applies before the execution, and one
that applies after.  The two functions must respect the relation
$\Phi(\state) \ge n + \Phi'(\state')$ for all states $\state$ and
$\state'$ such that $(S, \state) \Downarrow_n \state'$.  Intuitively,
$\Phi(\state)$ must provide enough \emph{potential} for both, paying
for resource cost of the computation and paying for the potential
$\Phi'(\state')$ on the resulting state $\state'$. That way, if
$(\state, S_1) \Downarrow_n \state'$ and $(\state', S_2) \Downarrow_m
\state''$, we get $\Phi(\state) \ge n + \Phi'(\state')$ and
$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed to
$\Phi(\state) \ge (n + m) + \Phi''(\state'')$.  Note that the initial
potential function $\Phi$ provides an upper bound on the resource
consumption of the whole program.  What we have observed is that, if
we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state')
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

@@

It is also possible to incorporate boolean preconditions into the
bounds to express that the bound is only valid for a certain class of
inputs.  To this end, we allow the potential function $\Phi$ in the
pre and post conditions to take non-negative number or $\infty$
(infinity) as values.  Infinity, plays the same role as
$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
into potential functions by mapping $\mathit{false}$ to $\infty$ and
$\mathit{true}$ to $0$.

\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}
${\color{blue}%
  \{ \#_1(a) + 2k \}}$
while (k > 0) {
  x=0;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  while (x < N && a[x] == 1) {
    ${\color{blue}%
      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
    a[x]=0;
    ${\color{blue}%
      \{ \#_1(a) + 2k + 1 \}}$
    tick(1);
    ${\color{blue}%
      \{ \#_1(a) + 2k \}}$
    x++; }
  ${\color{blue}%
    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
  if (x < N) {
    ${\color{blue}%
      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
    a[x]=1;
    ${\color{blue}%
      \{ \#_1(a) + 2(k-1) + 1 \}}$
    tick(1); }
  ${\color{blue}%
    \{ \#_1(a) + 2(k-1) \}}$
  k--;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
} ${\color{blue} \{ \#_1(a) \}}$

@@

\caption{Example derivation using amortized reasoning.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \lstinline{tick(n)} and
  0 to all other operations.
  }
\label{fig:xmplinc}
\end{figure}

We can use the logic to prove a concrete bound for a given resource
metric, or a bound that is parametric in a set of resource metrics.
For example, it is possible to derive the following quantitative
triple for a binary-search function \code{bsearch} that holds for all
stack metrics $M$.  
%
\begin{center}
${ \{ (Z = \log_2(h-l)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$
\end{center}
%
A stack metric assigns cost $0$ to all evaluation steps except
function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
and after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.  In the pre- and postcondition, the logical
variable $Z$ is used to relate the size of the input with the
potential that is left after the function call.  The logical part $Z =
\log_2(h-l)$ of the precondition is $0$ if $Z =
\log_2(\state_0(h)-\state_0(l))$ in the initial state $\state_0$ and
$\infty$ otherwise.

For simplicity, assume for the remainder of this section that we are
interested in bounding the number of \emph{ticks} in a program.  In
other words, we use a resource metric that assigns cost $n$ to the
function call \code{tick(n)} and cost $0$ to all other operations.
The argument $n$ is an integer and a negative number means that $-n$
resources are returned.

@@

@@

\begin{figure}[t]
  \centering
\begin{lstlisting}[mathescape]
${\color{blue}%
  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
while (x+K<=y) {
  ${\color{blue}%
  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  x=x+K; 
  ${\color{blue}%
  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  tick(T);
  ${\color{blue}%
  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
}
${\color{blue}%
 \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
\end{lstlisting}
  \caption{Automatic derivation of a tight bound on the number of
    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
    $T>0$ are not program variables but denote a concrete constants in
    the program.  The reasoning works uniformly for all such $K$ and
    $T$.}
  \label{fig:ex1}
\end{figure}
%
\newlength{\progwidth}

  \caption{Derivations of bounds on the number of ticks for
    challenging examples.  In the potential functions, we only mention
    the non-zero terms and in the logical context $\Gamma$ we only
    mention assertions that we actually use in the reasoning. In
    Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]|
    {+} 0.05|[0,y]|$.}
  \label{fig:ex_list}
\end{figure*}
%
%
More formally, we develop (\pref{sec:AAA}) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables that are currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations variables.  It is used to determine at the
assignment if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ is very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.

>>>>

% Comment for anonymization
% Also look for "SHORT" comments

%\def\fullversion{}

\documentclass[nocopyrightspace,preprint]{sigplanconf}

@@

%SHORT
\usepackage[normal,belowskip=-11pt,aboveskip=1pt]{caption} %Remove extra space around captiosn
\DeclareCaptionFormat{myformat}{#1#2#3\vspace{-4pt}\hrulefill}
\captionsetup[figure]{format=myformat}
\renewcommand{\captionlabelfont}{\bf}
%SHORT

@@

\input{definitions}
\begin{document}

%SHORT
\lstset{basicstyle=\tt\scriptsize}
\abovedisplayskip=3pt
\belowdisplayskip=3pt
\abovedisplayshortskip=3pt
\belowdisplayshortskip=3pt
%SHORT

@@

In this section, we informally introduce the quantitative program
logic and the automatic amortized analysis for for Clight programs.
% The purpose of this section is to build some intuition that should
% make the following sections easy to read.

@@

To obtain a compositional analysis we also have to take into account
the state resulting from a program's execution.  We thus use two
potential functions, one that applies before the execution, and one
that applies after.  The two functions must respect the relation
$\Phi(\state) \ge n + \Phi'(\state')$ for all states $\state$ and
$\state'$ such that $(S, \state) \Downarrow_n \state'$.  Intuitively,
$\Phi(\state)$ must provide enough \emph{potential} for both, paying
for resource cost of the computation and paying for the potential
$\Phi'(\state')$ on the resulting state $\state'$. That way, if
$(\state, S_1) \Downarrow_n \state'$ and $(\state', S_2) \Downarrow_m
\state''$, we get $\Phi(\state) \ge n + \Phi'(\state')$ and
$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed to
$\Phi(\state) \ge (n + m) + \Phi''(\state'')$.  Note that the initial
potential function $\Phi$ provides an upper bound on the resource
consumption of the whole program.  What we have observed is that, if
we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state')
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

@@

It is also possible to incorporate boolean preconditions into the
bounds to express that the bound is only valid for a certain class of
inputs.  To this end, we allow the potential function $\Phi$ in the
pre and post conditions to take non-negative number or $\infty$
(infinity) as values.  Infinity, plays the same role as
$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
into potential functions by mapping $\mathit{false}$ to $\infty$ and
$\mathit{true}$ to $0$.
%
\begin{figure}
\center
\begin{minipage}[b]{0.8\linewidth}
\begin{lstlisting}
${\color{blue}%
  \{ \#_1(a) + 2k \}}$
while (k > 0) {
  x=0;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
  while (x < N && a[x] == 1) {
    ${\color{blue}%
      \{ (a[x] = 1) + \#_1(a) + 2k \}}$
    a[x]=0;
    ${\color{blue}%
      \{ \#_1(a) + 2k + 1 \}}$
    tick(1);
    ${\color{blue}%
      \{ \#_1(a) + 2k \}}$
    x++; }
  ${\color{blue}%
    \{ (x \ge N \lor a[x] = 0) + \#_1(a) + 2k \}}$
  if (x < N) {
    ${\color{blue}%
      \{ (a[x] = 0) + \#_1(a) + 1 + 2(k-1) + 1 \}}$
    a[x]=1;
    ${\color{blue}%
      \{ \#_1(a) + 2(k-1) + 1 \}}$
    tick(1); }
  ${\color{blue}%
    \{ \#_1(a) + 2(k-1) \}}$
  k--;
  ${\color{blue}%
    \{ \#_1(a) + 2k \}}$
} ${\color{blue} \{ \#_1(a) \}}$

@@

\caption{Example derivation using amortized reasoning.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \lstinline{tick(n)} and
  0 to all other operations.
  }
\label{fig:xmplinc}
\end{figure}
%
\iffull{
}{}
We can use the logic to prove a concrete bound for a given resource
metric, or a bound that is parametric in a set of resource metrics.
For example, it is possible to derive the following quantitative
triple for a binary-search function \code{bsearch} that holds for all
stack metrics $M$.  
%
\begin{center}
${ \{ (Z = \log_2(h-\ell)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$
\end{center}
%
A stack metric assigns cost $0$ to all evaluation steps except
function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
and after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.  In the pre- and postcondition, the logical
variable $Z$ is used to relate the size of the input with the
potential that is left after the function call.  The logical part $Z =
\log_2(h-\ell)$ of the precondition is $0$ if $Z =
\log_2(\state_0(h)-\state_0(\ell))$ in the initial state $\state_0$ and
$\infty$ otherwise.

For simplicity, assume for the remainder of this section that we are
interested in bounding the number of \emph{ticks} in a program.  Thus,
we use a resource metric that assigns cost $n$ to the function call
\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
is an integer and a negative number means that $-n$ resources are
returned.

@@

\begin{figure}[t]
  \centering
\begin{lstlisting}[mathescape]
${\color{blue}%
  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
while (x+K<=y) {
  ${\color{blue}%
  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  x=x+K; 
  ${\color{blue}%
  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  tick(T);
  ${\color{blue}%
  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
}
${\color{blue}%
 \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
\end{lstlisting}
  \caption{Automatic derivation of a tight bound on the number of
    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
    $T>0$ are not program variables but denote a concrete constants in
    the program.  The reasoning works uniformly for all such $K$ and
    $T$.}
  \label{fig:ex1}
\end{figure}

%
\newlength{\progwidth}

  \caption{Derivations of bounds on the number of ticks for
    challenging examples.  In the potential functions, we only mention
    the non-zero terms and in the logical context $\Gamma$ we only
    mention assertions that we actually use in the reasoning. In
    Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]|
    {+} 0.05|[0,y]|$.}
  \label{fig:ex_list}
\end{figure*}
%
%

@@

More formally, we develop (\pref{sec:AAA}) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables that are currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations variables.  It is used to determine at the
assignment if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ is very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.
